---
title: "Saltmarsh Habitat Classification Models"
author: "Alyssa Bueno"
date: "2025-12-01"
output: 
  pdf_document:
      highlight: tango
      keep_tex: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Users/alyssabueno/Desktop/marshbirdsoutput/round_12")

library(terra)
library(randomForest)
library(dplyr)
library(caret)
library(sf)
library(tidyr)
library(ggplot2)
```

# PART 1: Load and Process Training Data

```{r}
# File paths
raster_files <- c(
  "/Users/alyssabueno/Desktop/marshbirds_data/clips/clip_bx.tif",
  "/Users/alyssabueno/Desktop/marshbirds_data/clips/clip_li.tif",
  "/Users/alyssabueno/Desktop/marshbirds_data/clips/clip_gb.tif"
)

shp_files <- c(
  "/Users/alyssabueno/Desktop/marshbirds_data/training_polygons/training_polygons_bx.shp",
  "/Users/alyssabueno/Desktop/marshbirds_data/training_polygons/training_polygons_li.shp",
  "/Users/alyssabueno/Desktop/marshbirds_data/training_polygons/training_polygons_gb.shp"
)

rasters <- lapply(raster_files, rast)
polygons <- lapply(shp_files, vect)

# -------------------------------
# Extract from all training sites
# -------------------------------

all_extracted <- list()
pca_model <- NULL  # tbd to store the PCA model

for (i in 1:length(rasters)) {
  naip <- rasters[[i]]
  polygon <- polygons[[i]]
  
  names(naip) <- paste0("naip", 1:4)
  
  # Calculate all features
  ndvi <- (naip[[4]] - naip[[1]]) / (naip[[4]] + naip[[1]])
  names(ndvi) <- "ndvi"
  
  ndwi <- (naip[[4]] - naip[[2]]) / (naip[[4]] + naip[[2]])
  names(ndwi) <- "ndwi"
  
  brightness <- (naip[[1]] + naip[[2]] + naip[[3]] + naip[[4]]) / 4
  names(brightness) <- "brightness"
  
  # Create PCA model from first training dataset, then reuse
  if (is.null(pca_model)) {
    all_for_pca <- c(naip, ndvi)
    vals <- values(all_for_pca)
    vals <- vals[complete.cases(vals), ]
    pca_model <- prcomp(vals, center = TRUE, scale. = TRUE)
  }
  
  # Apply PCA
  pca_input <- c(naip, ndvi)
  pca <- predict(pca_input, pca_model, index = 1:5)
  names(pca) <- paste0("PCA", 1:5)
  
  # Full feature stack (all features calculated)
  full_stack <- c(naip, ndvi, ndwi, brightness, pca)
  
  # Extract
  extracted_i <- terra::extract(full_stack, polygon, df = TRUE)
  extracted_i$class <- polygon$class[extracted_i$ID]
  extracted_i$ID <- NULL
  
  all_extracted[[i]] <- extracted_i
  cat("Processed training site", i, "\n")
}

# Combine all sites
extracted <- bind_rows(all_extracted)
extracted_clean <- na.omit(extracted)

# Sample balanced training data
training_data_full <- extracted_clean %>%
  group_by(class) %>%
  sample_n(min(2000, n())) %>%
  ungroup() %>%
  mutate(class = factor(class))

cat("\nTraining samples per class:\n")
print(table(training_data_full$class))

# Save PCA model 
saveRDS(pca_model, "~/Desktop/marshbirds_data/models/pca_model.rds")
```

# PART 2: Load and Calculate Test Data

```{r}
test_raster_files <- c(
  "/Users/alyssabueno/Desktop/marshbirds_data/test_clips/test_clips_bx.tif",
  "/Users/alyssabueno/Desktop/marshbirds_data/test_clips/test_clips_li.tif",
  "/Users/alyssabueno/Desktop/marshbirds_data/test_clips/test_clips_gb.tif"
)

test_shp_files <- c(
  "/Users/alyssabueno/Desktop/marshbirds_data/test_polygons/test_polygons_bx.shp",
  "/Users/alyssabueno/Desktop/marshbirds_data/test_polygons/test_polygons_li.shp",
  "/Users/alyssabueno/Desktop/marshbirds_data/test_polygons/test_polygons_gb.shp"
)

test_rasters <- lapply(test_raster_files, rast)
test_polygons <- lapply(test_shp_files, vect)

test_extracted <- list()

for (i in 1:length(test_rasters)) {
  naip <- test_rasters[[i]]
  polygon <- test_polygons[[i]]
  
  names(naip) <- paste0("naip", 1:4)
  
  # Calculate features (using SAME pca_model from training)
  ndvi <- (naip[[4]] - naip[[1]]) / (naip[[4]] + naip[[1]])
  names(ndvi) <- "ndvi"
  
  ndwi <- (naip[[4]] - naip[[2]]) / (naip[[4]] + naip[[2]])
  names(ndwi) <- "ndwi"
  
  brightness <- (naip[[1]] + naip[[2]] + naip[[3]] + naip[[4]]) / 4
  names(brightness) <- "brightness"
  
  # Apply SAME PCA model
  pca_input <- c(naip, ndvi)
  pca <- predict(pca_input, pca_model, index = 1:5)
  names(pca) <- paste0("PCA", 1:5)
  
  # Full stack
  full_stack <- c(naip, ndvi, ndwi, brightness, pca)
  
  # Extract
  extracted_i <- terra::extract(full_stack, polygon, df = TRUE)
  extracted_i$class <- polygon$class[extracted_i$ID]
  extracted_i$ID <- NULL
  
  test_extracted[[i]] <- extracted_i
  cat("Processed test site", i, "\n")
}

extracted_test <- bind_rows(test_extracted)
test_data_full <- na.omit(extracted_test) %>%
  mutate(class = factor(class))

cat("\nTest samples per class:\n")
print(table(test_data_full$class))
```

# ITERATION 1: Full Model (NAIP + NDVI + NDWI + Brightness + PCA)

```{r}
cat("\n========== ITERATION 1: Full Model ==========\n")

# Use all columns from training_data_full
rf_model_1 <- randomForest(class ~ ., 
                           data = training_data_full, 
                           ntree = 500)

# Test
preds_1 <- predict(rf_model_1, newdata = test_data_full)
cat("Test Accuracy:", round(mean(preds_1 == test_data_full$class), 4), "\n\n")

print(rf_model_1)
print(confusionMatrix(preds_1, test_data_full$class))

# Variable importance
imp_1 <- as.data.frame(rf_model_1$importance)
imp_1$feature <- rownames(imp_1)

ggplot(imp_1, aes(x = reorder(feature, MeanDecreaseGini), y = MeanDecreaseGini)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Model 1: Full Model Variable Importance",
       x = "Feature", y = "Mean Decrease Gini") +
  theme_minimal()

# Save model
saveRDS(rf_model_1, "~/Desktop/marshbirds_data/rf_model_1_full.rds")
```

# ITERATION 2: NDWI + NDVI + PCA

```{r}
cat("\n========== ITERATION 2: NDWI + NDVI + PCA ==========\n")

# Select columns
train_2 <- training_data_full %>% 
  select(ndwi, ndvi, starts_with("PCA"), class)

test_2 <- test_data_full %>% 
  select(ndwi, ndvi, starts_with("PCA"), class)

# Train
rf_model_2 <- randomForest(class ~ ., 
                           data = train_2, 
                           ntree = 500)

# Test
preds_2 <- predict(rf_model_2, newdata = test_2)
cat("Test Accuracy:", round(mean(preds_2 == test_2$class), 4), "\n\n")

print(rf_model_2)
print(confusionMatrix(preds_2, test_2$class))

# Save model
saveRDS(rf_model_2, "~/Desktop/marshbirds_data/rf_model_2_ndwi_ndvi_pca.rds")
```

# ITERATION 3: NAIP + NDVI

```{r}
cat("\n========== ITERATION 3: NAIP + NDVI ==========\n")

# Select columns
train_3 <- training_data_full %>% 
  select(starts_with("naip"), ndvi, class)

test_3 <- test_data_full %>% 
  select(starts_with("naip"), ndvi, class)

# Train
rf_model_3 <- randomForest(class ~ ., 
                           data = train_3, 
                           ntree = 500)

# Test
preds_3 <- predict(rf_model_3, newdata = test_3)
cat("Test Accuracy:", round(mean(preds_3 == test_3$class), 4), "\n\n")

print(rf_model_3)
print(confusionMatrix(preds_3, test_3$class))

# Save model
saveRDS(rf_model_3, "~/Desktop/marshbirds_data/rf_model_3_naip_ndvi.rds")
```

# ITERATION 4: NAIP Only

```{r}
cat("\n========== ITERATION 4: NAIP Only ==========\n")

# Select columns
train_4 <- training_data_full %>% 
  select(starts_with("naip"), class)

test_4 <- test_data_full %>% 
  select(starts_with("naip"), class)

# Train
rf_model_4 <- randomForest(class ~ ., 
                           data = train_4, 
                           ntree = 500)

# Test
preds_4 <- predict(rf_model_4, newdata = test_4)
cat("Test Accuracy:", round(mean(preds_4 == test_4$class), 4), "\n\n")

print(rf_model_4)
print(confusionMatrix(preds_4, test_4$class))

# Save model
saveRDS(rf_model_4, "~/Desktop/marshbirds_data/rf_model_4_naip.rds")
```

# ITERATION 5: PCA Only

```{r}
cat("\n========== ITERATION 5: PCA Only ==========\n")

# Select columns
train_5 <- training_data_full %>% 
  select(starts_with("PCA"), class)

test_5 <- test_data_full %>% 
  select(starts_with("PCA"), class)

# Train
rf_model_5 <- randomForest(class ~ ., 
                           data = train_5, 
                           ntree = 500)

# Test
preds_5 <- predict(rf_model_5, newdata = test_5)
cat("Test Accuracy:", round(mean(preds_5 == test_5$class), 4), "\n\n")

print(rf_model_5)
print(confusionMatrix(preds_5, test_5$class))

# Save model
saveRDS(rf_model_5, "~/Desktop/marshbirds_data/rf_model_5_pca.rds")
```

# PART 3: Classify New Rasters (Choose Your Model)

```{r}
# Choose which model to use for predictions
CHOSEN_MODEL <- 1  # Change this to 1, 2, 3, 4, or 5

# Load the chosen model
if (CHOSEN_MODEL == 1) {
  rf_model <- rf_model_1
  feature_cols <- names(training_data_full)[names(training_data_full) != "class"]
} else if (CHOSEN_MODEL == 2) {
  rf_model <- rf_model_2
  feature_cols <- c("ndwi", "ndvi", paste0("PCA", 1:5))
} else if (CHOSEN_MODEL == 3) {
  rf_model <- rf_model_3
  feature_cols <- c(paste0("naip", 1:4), "ndvi")
} else if (CHOSEN_MODEL == 4) {
  rf_model <- rf_model_4
  feature_cols <- paste0("naip", 1:4)
} else if (CHOSEN_MODEL == 5) {
  rf_model <- rf_model_5
  feature_cols <- paste0("PCA", 1:5)
}

cat("Using Model", CHOSEN_MODEL, "\n")
cat("Features:", paste(feature_cols, collapse = ", "), "\n\n")

# List of prediction rasters
files <- list.files(
  "~/Desktop/marshbirds_data/pred_clips",
  pattern = "_raster_200\\.tif$",
  full.names = TRUE
)

out_dir <- "~/Desktop/marshbirdsoutput/round_11/"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

# Process each raster
for (f in files) {
  message("Processing: ", basename(f))
  
  naip <- rast(f)
  names(naip) <- paste0("naip", 1:4)
  
  # Calculate all features (same as training)
  ndvi <- (naip[[4]] - naip[[1]]) / (naip[[4]] + naip[[1]])
  names(ndvi) <- "ndvi"
  
  ndwi <- (naip[[4]] - naip[[2]]) / (naip[[4]] + naip[[2]])
  names(ndwi) <- "ndwi"
  
  brightness <- (naip[[1]] + naip[[2]] + naip[[3]] + naip[[4]]) / 4
  names(brightness) <- "brightness"
  
  # Apply PCA (using saved model)
  pca_input <- c(naip, ndvi)
  pca <- predict(pca_input, pca_model, index = 1:5)
  names(pca) <- paste0("PCA", 1:5)
  
  # Full stack
  full_stack <- c(naip, ndvi, ndwi, brightness, pca)
  
  # Select only the features needed for this model
  prediction_stack <- full_stack[[feature_cols]]
  
  # Predict
  classified <- predict(prediction_stack, rf_model, na.rm = TRUE)
  
  # Save
  base <- tools::file_path_sans_ext(basename(f))
  out_file <- file.path(out_dir, paste0(base, "_model", CHOSEN_MODEL, "_classified.tif"))
  writeRaster(classified, out_file, overwrite = TRUE)
  
  message("Saved: ", out_file, "\n")
}

message("✔ ALL DONE")
```

# Visualize Results

```{r}
# Custom colors for classes
colors <- c(
  "#a6d96a",  # hm  
  "#1a9641",  # lm  
  "#8c510a",  # md  
  "#3288bd",  # ow  
  "#fdae61",  # ph  
  "#969696",  # rd  
  "#762a83"   # up  
)

# Plot one of the classified rasters
example_file <- list.files(out_dir, pattern = "_classified\\.tif$", full.names = TRUE)[1]
classified <- rast(example_file)
plot(classified, col = colors, main = basename(example_file))
```

# Compare Model Performance

```{r}
# Summary table
results <- data.frame(
  Model = c("Full Model", "NDWI+NDVI+PCA", "NAIP+NDVI", "NAIP Only", "PCA Only"),
  Features = c(
    "All",
    "NDWI, NDVI, PCA1-5",
    "NAIP1-4, NDVI",
    "NAIP1-4",
    "PCA1-5"
  ),
  Accuracy = c(
    mean(preds_1 == test_data_full$class),
    mean(preds_2 == test_2$class),
    mean(preds_3 == test_3$class),
    mean(preds_4 == test_4$class),
    mean(preds_5 == test_5$class)
  ),
  OOB_Error = c(
    tail(rf_model_1$err.rate[, 1], 1),
    tail(rf_model_2$err.rate[, 1], 1),
    tail(rf_model_3$err.rate[, 1], 1),
    tail(rf_model_4$err.rate[, 1], 1),
    tail(rf_model_5$err.rate[, 1], 1)
  )
)

print(results)

# Plot comparison
ggplot(results, aes(x = reorder(Model, Accuracy), y = Accuracy)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = round(Accuracy, 3)), hjust = -0.1) +
  coord_flip() +
  ylim(0, 1) +
  labs(title = "Model Comparison", x = "Model", y = "Test Accuracy") +
  theme_minimal()
```

## Key Changes:

1. **PCA calculated once** from first training site, then reused everywhere
2. **All features calculated upfront** in training/test data
3. **Each iteration just selects columns** from the full dataset (no recalculation)
4. **Models are saved** so you can reload them later
5. **Single prediction loop** that works for any model (just change `CHOSEN_MODEL`)
6. **Clear comparison** at the end to see which model performs best

Now you can just change `CHOSEN_MODEL` to switch between iterations without rerunning everything!